Computing the bicharacter of huge subspaces of polynomials

### 2019-07-24 article/article.tex
Check verbs to use present everywhere.

	\section{Introduction}
	In this article, we present a new SageMath package to compute bicharacter of some multivariate polynomials spaces. 
	The development of this package started at the occasion of a collaboration with François Bergeron to study the character of the space $\harmonics_n^{(r)}$ of diagonal harmonic polynomials on $r$ sets of $n$ variables, as a graded bi-$\GLr \times \Sn$-module. It was then extended to compute other characters of polynomial spaces.

	The main achievement is a calculation of the character of the diagonal harmonic polynomials for $n=6$ \pauline{+ some cases for $n=7$} and $r=5$ (and therefore any $r$) in 45 minutes (and ~15Gb), which led to conjectures on its general value.

	This is a potentially huge calculation ($\harmonics_6^{(5)}$ is of dimension $5.8 \cdot10^5$), involving polynomials of degree up to $15$ in $30$ variables with thousands of terms. So we had to combine several strategies that each reduced the time and/or memory complexity by one or two orders of magnitude. As much as possible, the code was written to be modular and reusable.

@@

	The invariant ring $R^{\Sn}$ is known as the ring of multisymmetric functions, that is to say the elements of $R$ that remain unchanged under permutations of the rows of $X$. The co-invariant ring $R_{\Sn}$ is the quotient of $R$ by the ideal $\mathcal{I}_n = \langle {R^{\Sn^+}} \rangle$ of the multisymmetric functions without constant term. It inherits from $R$ a structure of graded $\GLr \times \Sn$ bi-module. \pauline{Add details about grading.}

@@

    In order to study those spaces, we will use representation theory and characters.  Intuitively, through representation theory, the purpose is to cut modules into irreducible submodules according to the group action. Those submodules are then represented by irreducible characters that encodes the group action on each submodule. For $\GLr$ and $\Sn$, irreducible characters can be indexed by partitions and encoded by Schur functions. Knowing the irreducible characters gives informations of the studied module such as the dimension.

    We will not enter into details on representation theory here but for more on this topic see \cite{FultonHarris1991}. Also, \cite{Sagan} is a classical reference for representation theory of the symmetric group and see \cite{Green2007} for more on the polynomial representations of the general linear group. \\

   	The characters of the harmonic and diagonal harmonic polynomials spaces have been studied a lot until now. For example, it is known that, when r=1, the action of $GL_1$ is trivial and $R_{\Sn}$ carries the usual graded representation of $\Sn$. Its character is given by a Hall-Littlewood polynomial.

@@

	Throughout this article, we will write the spaces and their the character the same way, except if there might be a confusion, then the variables will be added to the character and we will write $\mathcal{H}_n^{(r)}(\bm{q};\bm{w})$ for the character of $\mathcal{H}_n^{(r)}$,  with $\bm{q}$ the variables used to express the $\GLr$ part of the character and $\bm{w}$ for the $\Sn$ part.\\

@@

	So we took the classical route of realizing the coinvariant ring as the space $\harmonics_n^{(r)}$ of diagonal harmonic polynomials.
	This is the space of those polynomials that satisfy the equations given by the multisymmetric functions seen as differential operators.

	Solving those differential equations would not be tractable either. So instead we used operators, partial derivatives and polarization operators defined in the previous section. \\

@@

	However, since we use the Young symmetrizer to construct $\harmonics_{n}^{(r)}$, for all $v \in \harmonics_{n}^{(r)}$, $v$ is anti-symmetric with respect to the column stabilizer $Q_\mu$, with $\mu$ the type of the isotypic component of $v$. Indeed, when projecting a polynomial $p$ on its isotypic component using the Young idempotent $c_\mu$, by definition $p.a_\mu$ is symmetric with respect to the row preserving permutations of $\mu$ and then $v = p.a_\mu.b_\mu$ is anti-symmetric with respect to the column preserving permutations of $\mu$. 
	It's therefore sufficient to keep one representative monomial per $S_P$ orbit \pauline{$S_P$ orbit?} (and do some appropriate straightening whenever a new monomial is created). We chose to keep the smallest representative monomial in lexicographic order and call the result the \emph{normal reduced form} of $v$. \\

>>>>

	\section{Introduction}
	In this article, we present a new SageMath package to compute bicharacter of some multivariate polynomials spaces. 
	The development of this package started at the occasion of a collaboration with François Bergeron to study the character of the space $\harmonics_n^{(r)}$ of diagonal harmonic polynomials on $r$ sets of $n$ variables, as a graded bi-$\GLr \times \Sn$-module. It is then extended to compute other characters of polynomial spaces.

	The main achievement is a calculation of the character of the diagonal harmonic polynomials for $n=6$ and $r=5$ (and therefore any $r$) in 45 minutes (and ~15Gb), which led to conjectures on its general value.

	This is a potentially huge calculation ($\harmonics_6^{(5)}$ is of dimension $5.8 \cdot10^5$), involving polynomials of degree up to $15$ in $30$ variables with thousands of terms. So we have to combine several strategies that each reduced the time and/or memory complexity by one or two orders of magnitude. As much as possible, the code is written to be modular and reusable.

@@

	The invariant ring $R^{\Sn}$ is known as the ring of multisymmetric functions, that is to say the elements of $R$ that remain unchanged under permutations of the rows of $X$. The co-invariant ring $R_{\Sn}$ is the quotient of $R$ by the ideal $\mathcal{I}_n = \langle {R^{\Sn^+}} \rangle$ of the multisymmetric functions without constant term. It inherits from R a structure of graded $\GLr \times \Sn$ bi-module. The degree of a polynomial in $R_{\Sn}$ is defined to be the sum of the degrees in each set of variables.

@@

    In order to study those spaces, we use representation theory and characters.  Intuitively, we want to decompose modules into irreducible submodules according to the group action. Those submodules are then represented by irreducible characters that encode the group action on each submodule. For $\GLr$ and $\Sn$, irreducible characters can be indexed by partitions and encoded by Schur functions. Knowing the irreducible characters gives information on the studied module such as the dimension.

    We don't enter into details on representation theory here but for more on this topic see \cite{FultonHarris1991}. Also, \cite{Sagan} is a classical reference for representation theory of the symmetric group and \cite{Green2007} is about polynomial representations of the general linear group. \\

   	The characters of the harmonic and diagonal harmonic polynomial spaces have been studied a lot until now. For example, it is known that, when r=1, the action of $GL_1$ is trivial and $R_{\Sn}$ carries the usual graded representation of $\Sn$. Its character is given by a Hall-Littlewood polynomial.

@@

	Throughout this article, we write the spaces and their the character the same way, except if there might be a confusion, then the variables are added to the character and we write $\mathcal{H}_n^{(r)}(\bm{q};\bm{w})$ for the character of $\mathcal{H}_n^{(r)}$,  with $\bm{q}$ the variables used to express the $\GLr$ part of the character and $\bm{w}$ for the $\Sn$ part.\\

@@

	So we take the classical route of realizing the coinvariant ring as the space $\harmonics_n^{(r)}$ of diagonal harmonic polynomials.
	This is the space of those polynomials that satisfy the equations given by the multisymmetric functions seen as differential operators.

	Solving those differential equations would not be tractable either. So instead we use partial derivatives and polarization operators defined in the previous section. \\

@@

	However, since we use the Young symmetrizer to construct $\harmonics_{n}^{(r)}$, for all $v \in \harmonics_{n}^{(r)}$, $v$ is anti-symmetric with respect to the column stabilizer $Q_\mu$, with $\mu$ the type of the isotypic component of $v$. Indeed, when projecting a polynomial $p$ on its isotypic component using the Young idempotent $c_\mu$, by definition $p.a_\mu$ is symmetric with respect to the row preserving permutations of $\mu$ and then $v = p.a_\mu.b_\mu$ is anti-symmetric with respect to the column preserving permutations of $\mu$. 
	It is therefore sufficient to keep one representative monomial per $S_P$ orbit \pauline{$S_P$ orbit?} (and do some appropriate straightening whenever a new monomial is created). We chose to keep the smallest representative monomial in lexicographic order and call the result the \emph{normal reduced form} of $v$. \\
